---
title: "Exponential distribution in R & comparison with the Central Limit Theorem"
author: "Twesigomwe David"
date: "November 28, 2017"
---
###Overview:
This part of the report covers the investigation of the exponential distribution in R in comparison with the centra limit theorem (CLT). The CLT states that the distribution of properly normalized averages of iid variables becomes that of a standard normal as the sample size, n increases. The exponential distribution was generated by simulation.

We know that for the exponential distribution;

the theoretical mean = theoretical standard deviation = 1/0.2 = 5

Variance = (1/0.2)**2 = 25

```{r, echo=FALSE}
theoretical_mean<-5
theoretical_variance<-25
```

###### Consider the disribution of 1000 random exponentials (distr_1)

```{r, echo=TRUE} 
distr_1<-(rexp(1000, .2))
sample_mean<-mean(distr_1)
sample_variance<-var(distr_1)
hist(distr_1, main = "Histogram showing rexp(1000, .2)")
legend("topright", legend = c(paste("Sample mean =", round(sample_mean, 2)), paste("sample variance = ", round(sample_variance, 2))))
```

The sample mean and variance for the 1000 random exponentials are consistent respectively with the theoretical mean and theoretical variance of an exponential distribution for 1000 simulations.

###### Now consider the distribution of 1000 averages of 40 exponentials (avs)

```{r}
avs = NULL 
for (i in 1:1000) avs = c(avs, mean(rexp(40, .2)))
a<-mean(avs)
b<-var(avs)
hist(avs, main = "Distribution of 1000 averages of 40 exponentials")
abline(v=a, col="red")
abline(v=theoretical_mean, col="blue")
legend("topright", legend = c(paste("Sample mean(red) =", round(a, 2)), paste("Theoretical mean(blue) = ", (theoretical_mean)), paste("Variance =", round(b, 2))))
```

This distribution is far more Gaussian compared to the original exponential distribution. It has a mean consistent with the theoretical mean. However the variance is much smaller since the distribution is now more Gaussian, reflecting the effect of having larger n as a result of the simulation. 


######In fact when we plot a distribution of 1000 normalized averages of 40 exponentials(mns);
```{r, echo=TRUE}
mns = NULL
for (i in 1:1000) mns = c(mns, (mean(rexp(40, .2))-5)/(5/sqrt(40)))
d<-mean(mns)
e<-var(mns)
hist(mns, prob=TRUE, density = 10, ylab = "Density", main = "Distribution of 1000 Normalized averages of 40 exponentials")
abline(v=d, col="blue")
lines(density(mns), col = "black")
legend("topright", legend = c(paste("mean(blue) =", round(d, 1)), paste("variance = ", round(e, 2))))
```

The distribution obtained approximates to a bell-shape and it is close to that of a standard normal with mean 0, and variance 1 as shown in the figure, thus conforming to the Central Limit Theorem.


